---
title: "Implementation Multiple Time Series Analysis"
author: "Jacopo Lussetti"
date: "2025-04-02"
header-includes:
  - \usepackage{amsmath}
  - \DeclareMathOperator{\VEC}{vec}
  - \usepackage{bm}  # Bold math symbols
output:
  html_document: default
  pdf_document: default
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
#install required packages
packages <- c("MASS", "zoo", "stats", "rockchalk", "glmnet", "vars", "knitr")

for (pckg in packages) {
  if (!require(pckg, character.only = TRUE)) {
    install.packages(pckg)
    library(pckg, character.only = TRUE)
  } else {
    library(pckg, character.only = TRUE)
  }
}



```   

## Companion Matrix function 


```{r}
comp_mtrx <- function(AA){
    ## AA is a K x Kp matrix, so we are able to derive p in the following way
    K <- nrow(AA)
    Kp <- ncol(AA)      
    p <- Kp/K
    
    # Create the empty companion matrix Kp x Kp
    C <- matrix(0, nrow=Kp, ncol=Kp)

    C[1:K,] <- AA
    
    # Add ones on the K-th sub-diagonal
    if (p>1)
        C[(K+1):Kp, 1:(Kp-K)] <- diag(Kp - K)
    return(C)
}
```

## Autocovariance function

Equation (2.1.39) page 29 represents the formula to compute autocovariances of 
a \textit{stable} VAR(p) Process. First and foremost we will then have to evaluate
whether the process is stable 

### Stability check 

A VAR(p) is a \textbf{stable process} if the condition (2.1.9.) holds: 

$$
det(I- \mathbf{A}z) \neq 0 \ if \ |z|<1 \ \text{where z are eigenvalues for A}
$$

```{r}
  var_roots<-function(AA){
    if(nrow(AA)==ncol(AA)){  # matrix is squared
      C<-AA  
    }else{
      C<-comp_mtrx(AA) # transform form compact matrix into companion matrix 
    }
    eig<-eigen(C)$values
    return(eig) 
  }
  
```

### Formula
After evaluating the conditions for stability, we proceed then defining the 
formula to compute contrivances. Th


\begin{equation}
  \VEC \ \Gamma_Y(0)= (I_{(Kp)^2}- \mathbf{A} \otimes \mathbf{A})^{-1} \VEC \ \Sigma_U
\end{equation}


```{r}

autocov_fun<-function(A, Sigma_u,p=1){ # A for high-order var is combined matrix,
  K<-nrow(Sigma_u) 
  Kp<-K * p
  #for var(1) is just A1
 if(p>1){
   #compute companion
   A<- comp_mtrx(A)
   #extend original sigma_u
  Sigma_U<-matrix(0, nrow=Kp, ncol=Kp) 
  Sigma_U[1:K, 1:K]<-Sigma_u
   
 }else{
   Sigma_U<-Sigma_u
 } 
 # compute the Kronecker product
  I<-diag(1, Kp^2)
  #compute vectorised Sigma_U
  vec_Sigma_U<-as.vector(Sigma_U)
  # compute the Autocovariance function
  vec_gamma_0<-solve(I - kronecker(A, A)) %*% vec_Sigma_U
  # reshape the result into a matrix
  Gamma_Y_0<-matrix(vec_gamma_0, nrow=Kp, ncol=Kp)
  
  return(Gamma_Y_0)  
}
```
## Equilibrium Points

Equilibrium points are defined by formula 2.1.10 at page 16

\begin{equation}
  \mathbf{\mu}:= E(Y_t)= (I_{Kp}-\mathbf{A})^{-1} \mathbf{\nu}
\end{equation}

```{r}
equilibrium<-function(A, nu){
  #check stability condition
  eig<-var_roots(A)
  if(any(Mod(eig)>=1)){   
      stop("Trajectories are not stable")
  }
  Kp<-nrow(A)
  I_Kp<-diag(1,Kp)
  values<-solve(I_Kp-A) %*% nu
  return(values)
}
```



## VAR(p) model

For this case we just consider the function simulating the trajectories, whereas
equilibrium and autocovariance functions are treated separately. 

```{r}
  var_sim <- function(AA, nu, Sigma_u, nSteps, y0) {
    K <- nrow(Sigma_u)
    Kp <- ncol(AA)
    p <- Kp/K
        
    if (p > 1) {
        C <- comp_mtrx(AA) # form the  companion matrix of the var(p) process
    } else {
        C <- AA  
    }
    y_t <- matrix(0, nrow =  nSteps, ncol=Kp) #trajectories matrix nSteps x Kp
    y_t[1, 1:Kp] <- y0 #add initial value to initiate the simulation
    noise <- mvrnorm(n = nSteps, mu = rep(0, K), Sigma = Sigma_u) #assuming that 
    #residuals follow a multivariate normal distribution    
    
    for (t in 2:nSteps) {
        y_t[t, ] <- C %*% y_t[t-1, ]
        y_t[t, 1:K] <- y_t[t, 1:K] + nu + noise[t,]
    }
    
    y_t <- zoo(y_t[,1:K], 1:nSteps)  
    return(y_t)
  }
```



# Estimates

## Estimates of coefficents
We first define a formula to compute Z values, which are key parameters for 
both estimating coefficient \& auto-correlation matrix. 

```{r}


par_estimate <- function(y_t, p=1) {
  nObs <- nrow(y_t)  # Number of observations
  K <- ncol(y_t)  # Number of variables
  T <- nObs - p  # Number of usable observations

  # y 
  Y <- y_t[(p + 1):nObs, ]  # T x K matrix

  # Z
  Z <- matrix(1, nrow = T, ncol = (K * p + 1))  # Intercept + lagged values

  for (i in 1:p) {
    col_start <- 2 + (i - 1) * K
    col_end <- 1 + i * K
    Z[, col_start:col_end] <- y_t[(p + 1 - i):(nObs - i), ]
  }

  # Estimate coefficients using OLS: B_hat = (Z'Z)^(-1) Z'Y
  B_hat <- solve(t(Z) %*% Z) %*% t(Z) %*% Y  # (K*p + 1) x K matrix

  return(list(
    Y = Y,
    Z = Z,
    B_hat = B_hat,  # Estimated VAR parameters
    T = T
  ))
}

```

```{r}
estimator<-function(Y, Z, p=1, method = c("standard", "qr", "lsfit")) {

    # Estimator
     method <- match.arg(method)
     if(method == "standard"){
        B_hat <- solve(t(Z) %*% Z, t(Z) %*% Y)
        B_hat <- t(B_hat)
     } else if(method == "qr"){
        qr_decomp <- qr(Z)  
        B_hat <- qr.coef(qr_decomp, Y)
     } else if(method == "lsfit"){
       fit <- lsfit(Z, Y) 
        B_hat <- fit$coef  
     } else {
        stop("Unknown method")
     }
    return(list(nu=B_hat[,1], AA=B_hat[,-1]))
  }

```


## Estimates of the autocovariance function

```{r}
est_autocov <- function(y_t, Y, Z, T, p=1){
    K <- ncol(y_t)
    Kp <- K * p
    I_t <- diag(T)

    # QR decomposition of Z to avoid singularity issues
    qr_decomp <- qr(Z)
    Q <- qr.Q(qr_decomp)
    P_Z <- Q %*% t(Q)  # Projection matrix

    # Compute bias-corrected covariance
    bias_sigma <- 1/T * t(Y) %*% (I_t - P_Z) %*% Y

    # Degrees of freedom correction
    d.f. <- T / (T - Kp - 1)
    unbiased <- d.f. * bias_sigma  # Corrected covariance estimate

    return(unbiased)
}

```
  
# Test A


```{r, fig.width=7, fig.height=5}
set.seed(123)
  p<-1
  A<-matrix(c(0.5, 0.1, 0., 0., 0.1, 0.2, 0., 0.3, 0.3), nrow=3, ncol=3)
  nu<-matrix(c(0.05, 0.02, 0.04), nrow=3)
  Sigma_u <- matrix(c(1.0, 0.2, 0.3, 0.2, 2.0, 0.5, 0.3, 0.5, 3.0), nrow = 3, ncol = 3)
  nSteps <- 200
  y0 <- matrix(c(0.1, -0.1, 0.2),ncol=ncol(A))

#compute trajectories
y_t_a<-var_sim(A, nu, Sigma_u, nSteps, y0)  
plot(y_t_a, main = "VAR(1) Simulation", xlab = "Time", ylab = "y values", col = 1:3, lty = 1)
```
## Multivariate Least Squares Estimators


```{r}
#estimate parameters
par_A<-par_estimate(y_t_a)
#estimate coefficient 

test_A<-estimator(par_A$Y, par_A$Z)
auto_cov_A<-est_autocov(y_t_a, par_A$Y, par_A$Z, par_A$T)

```
Now we will compare original input for the simulation \& estimated values

Coefficient Matrix
```{r}
A_true <- A  
A_est <- test_A$AA 

diff_A <- A_true - A_est  
print(diff_A)

```
Intercept Matrix
```{r}
nu_true <- nu  # True nu from the simulation
nu_est <- test_A$nu  # Estimated nu

diff_nu <- nu_true - nu_est  # Compute the difference
print(diff_nu)
```
Covariance Matrix

```{r}
Sigma_u_true <- Sigma_u  # True covariance matrix
Sigma_u_est <- auto_cov_A  # Estimated autocovariance

diff_Sigma_u <- Sigma_u_true - Sigma_u_est  # Compute the difference
print(diff_Sigma_u)

```
# Test B

Kp-dimensional representation for VAR(p) is defined as following:

\begin{align*}
Y_t &= \nu + \mathbf{A} Y_{t-1} + U_t \\
\text{where} \quad
Y_t &:= 
\begin{bmatrix}
y_t \\
y_{t-1} \\
\vdots \\
y_{t-p+1}
\end{bmatrix}, \quad
\nu :=
\begin{bmatrix}
\nu \\
0 \\
\text{($K p \times 1$ zeros)}
\end{bmatrix}, \\
\mathbf{A} &:=
\begin{bmatrix}
A_1 & A_2 & \ldots & A_{p-1} & A_p \\
I_K & 0 & \ldots & 0 & 0 \\
0 & I_K & \ldots & 0 & 0 \\
\vdots & & \ddots & \vdots & \vdots \\
0 & 0 & \ldots & I_K & 0
\end{bmatrix}, \quad
U_t :=
\begin{bmatrix}
u_t \\
0 \\
\vdots \\
0
\end{bmatrix}
\end{align*}

We first start with the simulation
```{r, fig.width=7, fig.height=5}
# Define parameters for Test B
set.seed(123)
p_B <- 2  # Number of lags for VAR(p)
nSteps_B <- 100
K_B <- 2  # Number of variables (size of y_t)
Sigma_u_B <- matrix(c(2, 0.3, 0.3, 3), nrow = 2, ncol = 2)
y0_B <- c(0.5, 2, 1, 5) 
nu_int_B <- matrix(c(0.5, 0.9), nrow = 2)

A_1 <- matrix(c(0.5, 0.4, 0.1, 0.5), nrow = 2, ncol = 2)
A_2 <- matrix(c(0, 0.25, 0, 0), nrow = 2, ncol = 2)
AA <- cbind(A_1, A_2)

# Simulate time series for Test B
y_t_B <- var_sim(AA, nu_int_B, Sigma_u_B, nSteps_B, y0_B)
plot(y_t_B, main = "VAR(2) Simulation", xlab = "Time", ylab = "y values", col = 1:2, lty = 1)

```
and then estimates

```{r}
# Estimate the parameters for Test B
par_B <- par_estimate(y_t_B, p = p_B)

# Estimate coefficients
test_B <- estimator(par_B$Y, par_B$Z)

# Display estimated coefficients for Test B
A_est_B <- test_B$AA
nu_est_B <- test_B$nu

# Estimate autocovariance matrix for Test B
auto_cov_B <- est_autocov(y_t_B, par_B$Y, par_B$Z, par_B$T, p = p_B)

# Display the estimated autocovariance matrix
auto_cov_B

```
Coefficient Matrix

```{r}
# True and estimated coefficient matrix for Test B
A_true_B <- AA  # True coefficient matrix
diff_A_B <- A_true_B - A_est_B  # Difference between true and estimated A

print("Difference in coefficient matrix (A):")
print(diff_A_B)

```
Intercept Matrix

```{r}
# True and estimated intercept matrix for Test B
nu_true_B <- nu_int_B  # True intercept vector
diff_nu_B <- nu_true_B - nu_est_B  # Difference between true and estimated nu

print("Difference in intercept matrix (nu):")
print(diff_nu_B)

```
Covariance Matrix

```{r}
# True and estimated covariance matrix for Test B
Sigma_u_true_B <- Sigma_u_B  # True covariance matrix
diff_Sigma_u_B <- Sigma_u_true_B - auto_cov_B  # Difference between true and estimated covariance

print("Difference in covariance matrix (Sigma_u):")
print(diff_Sigma_u_B)


```

# VAR(1) simulation with sparese coefficient Matrix

\begin{enumerate}
  \item \textbf{VAR(1) Simulation:} \\
  Generate data from the process
  \[
    y_t = A y_{t-1} + u_t, \quad u_t \sim \mathcal{N}(0, \sigma^2 I).
  \]
  where A is a lower triangular matrix with sparsity on the top right. 
  \item \textbf{Estimation via LASSO:} \\
  Apply LASSO regression to estimate the autoregressive matrix $A$ from the simulated data through R funtion \textit{glmnet}.
  \item \textbf{Optimisation of the tuning parameter} \\
    Through cross-validation or information criterion, and we compare strategies
    \item  \textbf{Monte Carlo Repetition}
  \item \textbf{Evaluation of Errors:} \\
  compute Type I and Type II error rates in detecting zero vs.~nonzero coefficients.
  \item \textbf{Increase dimension}\\
  We try larger matrices by increasing K, and we observe how LASSO’s performance changes as dimension increases.
\end{enumerate}

## VAR(1) Simulation

```{r}
set.seed(1234)
stab_test <- function(kp, A, tol = 1e-8)
{
  if (!is.matrix(A) || nrow(A) != ncol(A)) {
    stop("The matrix is not square")
  }
  eig <- eigen(A, only.values = TRUE)$values  # computing the eigenvalues
  
  for (i in 1:length(eig)) {     
    if (Mod(eig[i]) >= 1 - tol) { # Mod also handles complex numbers
      return(FALSE)               # <-- fixed typo "returm"
    }
  }
  return(TRUE)
}


A_sim <- function(K, spar, sd, max_tries = 1000){
  tries <- 0 
  repeat{
    A <- matrix(0L, K, K)
    idx_up <- which(upper.tri(A))
    n_up   <- length(idx_up)

    if (n_up > 0) {
      A[idx_up] <- rnorm(n_up, mean = 0, sd = sd)   #fill the upper 
  #triangle with r.v. from a normal distribution
  #cancel a certain percentage of the elements
      nmiss <- round(n_up * spar)
      if (nmiss > 0) { # add sparsity
        zero_idx <- sample(idx_up, nmiss, replace = FALSE)
        A[zero_idx] <- 0
      }
    }

    # check whether the model is stable (INSIDE the repeat loop)
    if (stab_test(K, A) == TRUE) return(A)

    tries <- tries + 1
    if (tries >= max_tries) {
      stop("Could not generate a stable A within max_tries.")
    }
  }
}

#now we generate a stable upper triangle coef matrix with sparsity 
K <- 5 
sd_var<-0.1
A <- A_sim(K, spar = 0.3, sd = sd_var, max_tries = 10) # moderate sparsity, modest sd A
stab_test(K,A)
```
Now we are able to generate a VAR(1) from the function specified at page 3

```{r}

sigma_var<- diag(1,5,5)* sd_var
nu_var <- rep(1, K)  
y_0 <- rep(0, K)  
T_sim<-10000
var_1<- var_sim(AA=A,nu=nu_var,Sigma_u=sigma_var, nSteps=T_sim, y0=y_0 )
```


# Estimation via LASSO

In this step we will apply lasso regression by applying the packages \textit{glmnet}

```{r}


# Build lagged design (Z) and aligned response (Y)
Z <- coredata(lag(var_1, k = -1))   # predictors y_{t-1}
Y_full <- coredata(var_1)                # responses  y_t

# Drop first row to align (remove the NA created by lag)
Y <- Y_full[-1, , drop = FALSE]

# If univariate, coerce to matrix to allow 2D subsetting later
if (is.null(dim(Z))) Z <- matrix(Z, ncol = 1)
if (is.null(dim(Y))) Y <- matrix(Y, ncol = 1)

# Chronological split (70/30)
n <- nrow(Z)
cut_off <- round(0.70 * n)
X_train <- Z[1:cut_off, , drop = FALSE]
Y_train <- Y[1:cut_off, , drop = FALSE]
X_test  <- Z[(cut_off + 1):n, , drop = FALSE]
Y_test  <- Y[(cut_off + 1):n, , drop = FALSE]

# Manual CV over a simple lambda grid (avoid 0 exactly)
lambdas <- seq(0,1, 0.001)
val_error <- data.frame(lambda = lambdas, error = NA_real_)

for (i in seq_along(lambdas)) {
  lmbd <- lambdas[i]
  fit  <- glmnet(X_train, Y_train, family = "mgaussian", lambda = lmbd)
  pred <- predict(fit, newx = X_test, s = lmbd)[,,1]
  mse_per_series   <- colMeans((Y_test - pred)^2)
  val_error$error[i] <- mean(mse_per_series)
}

best_lambda_1 <- val_error$lambda[which.min(val_error$error)]
best_lambda_1
```
We try with default cross-validation from glmnet

```{r}
cv_fit <- cv.glmnet(
  x = as.matrix(Z),
  y = as.matrix(Y),
  family = "mgaussian"  # <-- tell glmnet it's multivariate
)

plot(cv_fit)
best_lambda_2 <- cv_fit$lambda.min
best_lambda_2
```
```{r}
fit_opt<-glmnet(
  x=as.matrix(Z),
  y = as.matrix(Y),
  family = "mgaussian",
  lambda = c(best_lambda_1,best_lambda_2)
)
A_hat_114<-coef(fit_opt, s=best_lambda_1)
A_hat_129<- coef(fit_opt, s = best_lambda_2)
#we merge estimates in unique A_hat
A_hat_matrix_1<-matrix(NA, nrow=K, ncol=K)
for(i in 1:5){
  ci <- as.matrix(A_hat_114[[i]])           # ith response
lag_rows <- rownames(ci) != "(Intercept)"  # drop intercept
A_hat_matrix_1[i, ] <- ci[lag_rows, 1]
}
A_hat_matrix_1

A_hat_matrix_2<-matrix(NA, nrow=K, ncol=K)
for(i in 1:5){
ci <- as.matrix(A_hat_129[[i]])           # ith response
lag_rows <- rownames(ci) != "(Intercept)"  # drop intercept
A_hat_matrix_2[i, ] <- ci[lag_rows, 1]
}
A_hat_matrix_2

```
Now we check 

\begin{center}
\begin{tabular}{c|c|c|c|c}
   Lasso Regression & Type I & Type II & Recall & Precision & F1
   $\text{Lasso}_{\lambda=0.002}$ 
\end{tabular}
\end{center}
```{r}

# --- Support function
tol <- 1e-3
S_true  <- abs(A)                 > tol
S_est_1 <- abs(A_hat_matrix_1)    > tol
S_est_2 <- abs(A_hat_matrix_2)    > tol

# --- Confusion counts 
TP_1 <- sum(S_true & S_est_1)
FN_1 <- sum(S_true & !S_est_1)
FP_1 <- sum(!S_true & S_est_1)
TN_1 <- sum(!S_true & !S_est_1)

TP_2 <- sum(S_true & S_est_2)
FN_2 <- sum(S_true & !S_est_2)
FP_2 <- sum(!S_true & S_est_2)
TN_2 <- sum(!S_true & !S_est_2)

# --- Rates (guard divisions) ---
FPR_1 <- if ((FP_1+TN_1)>0) FP_1/(FP_1+TN_1) else NA_real_
FPR_2 <- if ((FP_2+TN_2)>0) FP_2/(FP_2+TN_2) else NA_real_

FNR_1 <- if ((FN_1+TP_1)>0) FN_1/(FN_1+TP_1) else NA_real_
FNR_2 <- if ((FN_2+TP_2)>0) FN_2/(FN_2+TP_2) else NA_real_

Prec_1 <- if ((TP_1+FP_1)>0) TP_1/(TP_1+FP_1) else NA_real_
Prec_2 <- if ((TP_2+FP_2)>0) TP_2/(TP_2+FP_2) else NA_real_

Rec_1  <- if ((TP_1+FN_1)>0) TP_1/(TP_1+FN_1) else NA_real_
Rec_2  <- if ((TP_2+FN_2)>0) TP_2/(TP_2+FN_2) else NA_real_

f1_1 <- if (!is.na(Prec_1+Rec_1) && (Prec_1+Rec_1)>0) 2*Prec_1*Rec_1/(Prec_1+Rec_1) else NA_real_
f1_2 <- if (!is.na(Prec_2+Rec_2) && (Prec_2+Rec_2)>0) 2*Prec_2*Rec_2/(Prec_2+Rec_2) else NA_real_

# Result Tables
results <- data.frame(
  Model     = c(paste0("Lambda=", best_lambda_1), paste0("Lambda=",best_lambda_2)),
  TypeI     = c(FPR_1, FPR_2),
  TypeII    = c(FNR_1, FNR_2),
  Recall    = c(Rec_1, Rec_2),
  Precision = c(Prec_1, Prec_2),
  F1        = c(f1_1, f1_2)
)
results
```


## Monte Carlo Simulation
```{r}
#define functions that returns 
error_function <- function(tol, coef, tr_coef){  # coef = A_hat, tr_coef = A_true
  S_true <- abs(tr_coef) > tol
  S_est  <- abs(coef)    > tol

  TP <- sum(S_true & S_est)
  FN <- sum(S_true & !S_est)
  FP <- sum(!S_true & S_est)
  TN <- sum(!S_true & !S_est)

  FPR  <- FP/(FP+TN)
  FNR  <-  FN/(FN+TP)
  Prec <-TP/(TP+FP)
  Rec  <-  TP/(TP+FN)
  F1   <- 2*Prec*Rec/(Prec+Rec) 
  
  data.frame(TypeI = FPR, TypeII = FNR, Recall = Rec, Precision = Prec, F1 = F1)
}


MC_var_1 <- function(K, spar, sd, nu, max_tries = 1000, nrep = 100, base_seed = 123, 
                     T_sim = 1000, tol = 1e-3) {

  seeds <- base_seed + seq_len(nrep) - 1

  A_list <- vector("list", nrep)  # store true A per replication (if you want)
  results_comb <- data.frame(
    TypeI     = numeric(nrep),
    TypeII    = numeric(nrep),
    Recall    = numeric(nrep),
    Precision = numeric(nrep),
    F1        = numeric(nrep)
  )

  for (b in seq_len(nrep)) {
    set.seed(seeds[b])

    # 1) draw a stable sparse A (your A_sim needs K as first arg)
    A_true <- A_sim(K, spar = spar, sd = sd, max_tries = max_tries)

    # 2) simulate VAR(1)
    sigma_var <- diag(1, K, K)                 # noise covariance (identity)
    nu_var    <- rep(nu, length.out = K)       # use the 'nu' argument
    y_0       <- rep(0, K)

    var_1 <- var_sim(AA = A_true, nu = nu_var, Sigma_u = sigma_var, 
                     nSteps = T_sim, y0 = y_0)

    # 3) design matrices Z = y_{t-1}, Y = y_t  (drop NA instead of using lag)
    Y <- coredata(var_1)[-1, , drop = FALSE]
    Z <- coredata(var_1)[-nrow(var_1), , drop = FALSE]

    # 4) CV to pick lambda
    cv_fit <- cv.glmnet(x = as.matrix(Z), y = as.matrix(Y), family = "mgaussian")
    lam <- cv_fit$lambda.min

    # 5) fit at chosen lambda and reconstruct A_hat
    fit_best   <- glmnet(x = as.matrix(Z), y = as.matrix(Y), family = "mgaussian", lambda = lam)
    coef_list  <- coef(fit_best)                 # list length K
    A_est      <- matrix(NA_real_, K, K)

    for (i in 1:K) {
      ci <- as.matrix(coef_list[[i]])            # (Intercept + K lag terms)
      A_est[i, ] <- ci[rownames(ci) != "(Intercept)", 1]  # drop intercept
    }

    # 6) metrics for this replication
    res_b <- error_function(tol = tol, coef = A_est, tr_coef = A_true)  # NOTE: coef = A_hat, tr = A_true
    results_comb[b, ] <- res_b[1, ]

    # (optional) keep the true A
    A_list[[b]] <- A_true
  }

  # return everything you need
  list(
    results = results_comb,
    summary = data.frame(
      metric = names(results_comb),
      mean   = sapply(results_comb, mean, na.rm = TRUE),
      sd     = sapply(results_comb, sd,   na.rm = TRUE),
      se     = sapply(results_comb, function(x) sd(x, na.rm = TRUE)/sqrt(sum(!is.na(x))))
    ),
    A_list = A_list,
    seeds  = seeds
  )
}
out <- MC_var_1(K = 5, spar = 0.3, sd = 0.1, nu = 0, nrep = 1000, T_sim = 1000)
out$summary


```
# Increasing number of K

We first try with K=10
```{r}
out_10 <- MC_var_1(K = 10, spar = 0.3, sd = 0.1, nu = 0, nrep = 1000, T_sim = 1000)
out_10$summary

```

We first try with K=15
```{r}

out_15 <- MC_var_1(K = 15, spar = 0.3, sd = 0.1, nu = 0, nrep = 1000, T_sim = 1000)
out_15$summary
```

